\emph{MixFishSim}: highly resolved spatiotemporal simulations for exploring mixed fishery dynamics

### 2019-07-12 write_up/mixfishsim_ecolmod.tex
Which witch is which

\replaced[id=PD]{T}{In the simulation model, t}he habitat for each of the
populations \replaced[id=PD]{was}{is} generated \replaced[id=PD]{with}{through}
the \textit{RFSimulate} function of the \textit{RandomFields} R package
\citep{Schlater2015}, which simulates a Gaussian Random Field process given a
user defined error model and correlation structure. We define a stationary
habitat field and combine with a temporally dynamic thermal tolerance field to
imitate two key drivers of population dynamics. Each population
\replaced[id=PD]{was}{is} initialised at a single location, and subsequently
move\replaced[id=PD]{ed}{s} according to a probabilistic
distrib\added[id=PD]{u}tion based on habitat suitability (represented by the
normalised values from the GRFs), temperature and distance
from current cell\replaced[id=PD]{:}{.} 
\begin{equation}
	Pr(J | I) = \frac{e^{-\lambda * d_{IJ}} \cdot
		(Hab_{J,p}^2 \cdot Tol_{J,p,wk})}{\sum\limits_{c=1}^{C}e^{-\lambda * d} \cdot
		(Hab_{c,p}^2 \cdot Tol_{c,p,wk})}
\end{equation}
Where $d_{IJ}$ is the euclidean distance between cell $I$ and cell $J$,
$\lambda$ is a given rate of decay, $Hab_{J,p}^2$ is the squared index of
habitat suitability for cell $J$ and population $p$, with $Tol_{J,p,wk}$ the
temperature tolerance for cell $J$ by population $p$ in week $wk$ (see
below).\\

@@

\added[id=PD]{The final process results in a population structure and movement
	pattern unique to each species, with population movement occurring on a
	weekly basis. The decision to model population movement on a weekly
	timescale was to reflect that fish tend to aggregate in species
	specific locations which have been observed to last around one to two
	weeks \citep{Poos2007}.  Therefore this process approximated the
	demographic shifts in fish populations throughout a year with seasonal
	spawning patterns (e.g.  Figure \ref{S-fig:5}).}

@@

The fleet dynamics can be broadly categorised into three components; fleet
targeting - which determine\replaced[id=PD]{d}{s} the fleet catch efficiency
and preference towards a particular species; trip-level decisions, which
determine\added[id=PD]{d} the initial location to be fished at the beginning of
a trip; and within-trip decisions, determining movement from one fishing spot
to another within a trip. Together, these elements implement an explore-exploit
type strategy for individual vessels to maximise their catch from an unknown
resource distribution \cite{Bailey2018}. The decision to use an individual
based model for fishing vessels was taken because fishers are heterogeneous in
their location choice behaviour due to different objectives, risk preference
and targeting preference \citep{VanPutten2012a}. Therefore in the simulations
fleet dynamics are the productive of individual experiences rather than
pre-defined group dynamics.

@@

Several studies \citep[e.g.][]{Hutton2004, Tidd2012, Girardin2015} have
confirmed past activity and past catch rates are strong predictors of fishing
location choice. For this reason, the fleet dynamics sub-model
include\replaced[id=PD]{d}{s} a learning component, where a vessel's initial
fishing location in a trip \replaced[id=PD]{wa}{i}s based on selecting from
previously successful fishing locations. This \replaced[id=PD]{wa}{i}s achieved
by \added[id=PD]{calculating an expected revenue based on the catches from
	locations fished} in the preceding trip as well as the same month
periods in previous years and the travel costs from the port to the fishing
grounds, and choosing randomly from the top 75 \% of fishing events
\replaced[id=PD]{as defined by the expected profit, which has a seasonal
	component}{in value}.

@@

For our implementation of a random walk directional change is based on a
negatively correlated circular distribution where a favourable fishing ground
is likely to be ``fished back over" by the vessel returning in the direction it
came from.  The step length (i.e. the distance travelled from the current to
the next fishing location) is determined by \deleted[id=JJ]{relating} recent
fishing success, measured as the summed value of fish caught (revenue, $Rev$),
\begin{equation}
Rev = \sum_{p=1}^{P} \replaced[id=PD]{L}{C}_{p} \cdot Pr_{p} 
\end{equation}
where $\replaced[id=PD]{L}{C}_{p}$ is \replaced[id=PD]{landings}{catch} of a
population $p$, and $Pr_{p}$ price of a population. Here, when fishing is
successful vessels remain in a similar location and continue to exploit the
local fishing grounds. When unsuccessful, they move some distance away from
the current fishing location.  The movement distance retains some degree of
stochasticity, which can be controlled separately, but is determined by the
relationship: 
\begin{equation}
	StepL = e^{log(\beta_{1}) + log(\beta_{2}) -
		(log(\frac{\beta_{1}}{\beta_{3}}))} * Rev
\end{equation}\todo{\added[id=JJ]{So
			step length increases with increasingly gross
			revenue?}\added[id=PD]{No, the opposite}}
Where $\beta_{1}$, $\beta_{2}$ and $\beta_{3}$ are parameters determining the
shape of the step function in its relation to revenue, so that, a step from
(x1,y1) to (x2, y2) is defined by:
\begin{equation}
	\begin{split}
 (x2, y2) =  & x1 + StepL \cdot \cos (\frac{\pi \cdot Br}{180}), \\
             & y1 + StepL \cdot \sin (\frac{\pi \cdot Br}{180}) \\	
 with  \hspace{0.5cm}     & Br_{t-1} < 180, Br_{t} = 180 + \sim vm[(0,360), k] \\
 			  & Br_{t-1} > 180, Br_{t} = 180 - \sim vm[(0,360), k] \\
	\end{split}
\end{equation}
where $k$ the concentration parameter from the von \replaced[id=JJ]{M}{m}ises
distribution which we correlate with the revenue so that $k = (Rev + 1 /
RefRev) * max_{k}$, where $max_{k}$ is the maximum concentration value, $k$,
and $RefRev$ is parametrised as for $\beta_{3}$ in the step length function. A
realised example of the step length and turning angle relationships to revenue
can be seen at Figure \ref{S-fig:15}.

@@

We parametrised the simulation model for four populations with different
demographics; growth rates, natural mortality and recruitment functions (Table
\ref{tab:1}). Habitat preference (Figure \ref{S-fig:1}) and temperature
tolerances (Figures \ref{S-fig:3}, \ref{S-fig:4}) were unique to each
population resulting in differently weekly distribution patterns (Figures
\ref{S-fig:5}-\ref{S-fig:7}). In addition, each of the populations has two
defined spawning areas which result in the populations moving towards these
areas in pre-defined weeks (Figure \ref{S-fig:2}) with population-specific
movement rates (Table \ref{tab:1}). The individual habitat preferences and
thermal tolerances result in different spatial habitat use for each population
(Figure \ref{S-fig:9}) and consequently different seasonal exploitation
patterns (Fishing mortality in Figure \ref{S-fig:10}).

@@

With increasing probability throughout the simulation, fishing locations were
chosen based on experience of profitable catches built up in the same month
from previous years and from the previous trip. 'Profitable' in this context
was defined as the locations where the top 70 \% of expected profit would be
found given previous trips revenue and cost of movement to the new fishing
location.  This probability was based on a logistic sigmoid function with a
lower asymptote of 0 and upper asymptote of 0.95, and a growth rate which
ensures the upper asymptote (where decisions are mainly based on past
knowledge) is reached approximately halfway through the simulation.  \\

@@

The following steps are undertaken to determine closures:
\todo{\added[id=JJ]{Procedure unclear. Refer to symbols in methods section or
		switch order starting with description of data type
		etc..}\added[id=PD]{Yes, will redo}}
\begin{enumerate}
	\item Extract data source
	\item Aggregate according to desired spatial and temporal resolution
	\item Interpolate across entire area at desired resolution using simple
		kriging using the \emph{interp} function from the R package
		akima \citep{Akima2006}.
	\item Close area covering top 5 \% of catch rates 
\end{enumerate}
In total 28 closure scenarios were run which represent combinations of:

@@

Vessels from the same fleet (and therefore targeting preference) exploit
similar but slightly different fishing grounds depending on their own personal
experience during the explore phase of the fishery (\added[id=PD]{Figure}
\ref{S-fig:14}), which is the result of the randomness in the correlated random
walk step function, with distance moved during the exploitation phase and the
direction stochastically related to the revenue experienced on the fishing
ground (\replaced[id=PD]{Figure}{while}
\ref{S-fig:15}).\todo{\added[id=JJ]{Move some of the supplementary figures to
		the manuscript}}

@@

In order to answer this question we compare different spatial and temporal
aggregations of the 'real population' distributions to:
\begin{enumerate}[label=\alph*)]
	\item \textbf{fisheries-independent data:} the inferred population from
		a fixed-site sampling survey design as commonly used for
		fisheries monitoring purposes;
	\item \textbf{fisheries-dependent data:} the inferred population from
		our fleet model which includes fishery-induced sampling
		dynamics.
\end{enumerate}

@@

The trend in fishing mortality for each species show that in most cases the
fishery closure was successful in reducing fishing mortality on the species of
interest (species 3; Figure \ref{fig:3}), though interestingly the largest
reductions in fishing mortality happened immediately after the closures,
following which the fisheries "adapted" to the closures and fishing mortality
increased again somewhat. The exception to the success was the closures
implemented based on the coarsest spatial (20 x 20) and temporal resolution
(yearly) which were ineffective with all data sources. As expected, closures
based on the "known" population distribution were most effective, with
differing degrees of success using the commercial data. Fishing mortality rates
on the other species changed in different proportions, depending on whether the
displaced fishing effort moved to areas where the populations were found in
greater or lesser density. \\

A regression tree (using the R package REEMtree \citep{Sela2012}) highlights
that the factor most contributing to differences in fishing mortality before
and after the closure was the population (72 \% showing that the closures were
effective for population 3), followed by data resolution (21 \%), data type (7
\%) with the least important factor the timescale ($<$ 1 \%). In general the
finer the spatial resolution of the data used the greater reduction in fishing
mortality for population 3 after the closures (Figure \ref{fig:4}). The notable
outliers are the commercial data at the coarsest spatial resolution (20 x 20)
at a yearly and weekly timescale, where closures were nearly as effective as
the fine-scale resolution. In this case the closures were sufficiently large to
protect a core are of the habitat for the population, but this was achieved in
a fairly crude manner by closing a large area - including area where the
species was not found (Figure \ref{S-fig:17}) which may have consequences in
terms of restricting the fishery in a much larger area than necessary. \\

@@

%% Comparison to other models
Typically, simulation models which treat fish as individuals are focussed on
exploring the inter- and intra- specific interactions among fish populations
(e.g. OSMOSE \cite{Shin2004}) in order to understand how they vary over space
and time. Our focus was on understanding the strengths and limitations of
inference from catch data obtained through commercial fishing activity with
fleets exploiting multiple fish populations and realising catch distributions
which may differ from the underlying populations. As such, we favoured a
minimum realistic model of the fish populations \citep{Plaganyi2014}, while
incorporating detailed fishing dynamics which take account of different drivers
in a mechanistic way. In this way we take account of heterogeneity in fleet
dynamics due to different preferences and drivers similarly to other approaches
\citep{Fulton2011}, but at an individual vessel rather than fleet level. We do
not explicitly define fleets as rational profit maximisers at the outset, but
consider there are several stages to development of the fishery; information
gathering through search where the resource location is not known,  followed by
individual learnt behaviour of profitable locations.  This provides a realistic
model of how fishing patterns are established and maintained to exploit an
uncertain resource through an explore-exploit strategy \citep{Mangel1983,
	Bailey2018}.

@@

In addition, we have assumed that fishing vessels are not restricted by quota
and therefore discarding of species for which vessels have no quota or which
are unwanted is not taken into account. This is likely to be a significant
source of bias in any inference using commercial data and should also be
explored. \\

>>>>

\replaced[id=PD]{T}{In the simulation model, t}he habitat for each of the
populations \replaced[id=PD]{was}{is} generated \replaced[id=PD]{with}{through}
the \textit{RFSimulate} function of the \textit{RandomFields} R package
\citep{Schlater2015}, that simulates a Gaussian Random Field process given a
user defined error model and correlation structure. We define a stationary
habitat field and combine with a temporally dynamic thermal tolerance field to
imitate two key drivers of population dynamics. Each population
\replaced[id=PD]{was}{is} initialised at a single location, and subsequently
move\replaced[id=PD]{ed}{s} according to a probabilistic
distrib\added[id=PD]{u}tion based on habitat suitability (represented by the
normalised values from the GRFs), temperature and distance
from current cell\replaced[id=PD]{:}{.} 
\begin{equation}
	Pr(J | I) = \frac{e^{-\lambda * d_{IJ}} \cdot
		(Hab_{J,p}^2 \cdot Tol_{J,p,wk})}{\sum\limits_{c=1}^{C}e^{-\lambda * d} \cdot
		(Hab_{c,p}^2 \cdot Tol_{c,p,wk})}
\end{equation}
Where $d_{IJ}$ is the euclidean distance between cell $I$ and cell $J$,
$\lambda$ is a given rate of decay, $Hab_{J,p}^2$ is the squared index of
habitat suitability for cell $J$ and population $p$, with $Tol_{J,p,wk}$ the
temperature tolerance for cell $J$ by population $p$ in week $wk$ (see
below).\\

@@

\added[id=PD]{The final process results in a population structure and movement
	pattern unique to each species, with population movement occurring on a
	weekly basis. The decision to model population movement on a weekly
	timescale was to reflect that fish tend to aggregate in species
	specific locations that have been observed to last around one to two
	weeks \citep{Poos2007}.  Therefore this process approximated the
	demographic shifts in fish populations throughout a year with seasonal
	spawning patterns (e.g.  Figure \ref{S-fig:5}).}

@@

The fleet dynamics can be broadly categorised into three components; fleet
targeting - that determines\replaced[id=PD]{d}{s} the fleet catch efficiency
and preference towards a particular species; trip-level decisions, that 
determines\added[id=PD]{d} the initial location to be fished at the beginning of
a trip; and within-trip decisions, determining movement from one fishing spot
to another within a trip. Together, these elements implement an explore-exploit
type strategy for individual vessels to maximise their catch from an unknown
resource distribution \cite{Bailey2018}. The decision to use an individual
based model for fishing vessels was taken because fishers are heterogeneous in
their location choice behaviour due to different objectives, risk preference
and targeting preference \citep{VanPutten2012a}. Therefore in the simulations
fleet dynamics are the productive of individual experiences rather than
pre-defined group dynamics.

@@

Several studies \citep[e.g.][]{Hutton2004, Tidd2012, Girardin2015} have
confirmed past activity and past catch rates are strong predictors of fishing
location choice. For this reason, the fleet dynamics sub-model
include\replaced[id=PD]{d}{s} a learning component, where a vessel's initial
fishing location in a trip \replaced[id=PD]{wa}{i}s based on selecting from
previously successful fishing locations. This \replaced[id=PD]{wa}{i}s achieved
by \added[id=PD]{calculating an expected revenue based on the catches from
	locations fished} in the preceding trip as well as the same month
periods in previous years and the travel costs from the port to the fishing
grounds, and choosing randomly from the top 75 \% of fishing events
\replaced[id=PD]{as defined by the expected profit, that has a seasonal
	component}{in value}.

@@

For our implementation of a random walk directional change is based on a
negatively correlated circular distribution where a favourable fishing ground
is likely to be ``fished back over" by the vessel returning in the direction it
came from.  The step length (i.e. the distance travelled from the current to
the next fishing location) is determined by \deleted[id=JJ]{relating} recent
fishing success, measured as the summed value of fish caught (revenue, $Rev$),
\begin{equation}
Rev = \sum_{p=1}^{P} \replaced[id=PD]{L}{C}_{p} \cdot Pr_{p} 
\end{equation}
where $\replaced[id=PD]{L}{C}_{p}$ is \replaced[id=PD]{landings}{catch} of a
population $p$, and $Pr_{p}$ price of a population. Here, when fishing is
successful vessels remain in a similar location and continue to exploit the
local fishing grounds. When unsuccessful, they move some distance away from
the current fishing location.  The movement distance retains some degree of
stochasticity, that can be controlled separately, but is determined by the
relationship: 
\begin{equation}
	StepL = e^{log(\beta_{1}) + log(\beta_{2}) -
		(log(\frac{\beta_{1}}{\beta_{3}}))} * Rev
\end{equation}\todo{\added[id=JJ]{So
			step length increases with increasingly gross
			revenue?}\added[id=PD]{No, the opposite}}
Where $\beta_{1}$, $\beta_{2}$ and $\beta_{3}$ are parameters determining the
shape of the step function in its relation to revenue, so that, a step from
(x1,y1) to (x2, y2) is defined by:
\begin{equation}
	\begin{split}
 (x2, y2) =  & x1 + StepL \cdot \cos (\frac{\pi \cdot Br}{180}), \\
             & y1 + StepL \cdot \sin (\frac{\pi \cdot Br}{180}) \\	
 with  \hspace{0.5cm}     & Br_{t-1} < 180, Br_{t} = 180 + \sim vm[(0,360), k] \\
 			  & Br_{t-1} > 180, Br_{t} = 180 - \sim vm[(0,360), k] \\
	\end{split}
\end{equation}
where $k$ the concentration parameter from the von \replaced[id=JJ]{M}{m}ises
distribution that we correlate with the revenue so that $k = (Rev + 1 /
RefRev) * max_{k}$, where $max_{k}$ is the maximum concentration value, $k$,
and $RefRev$ is parametrised as for $\beta_{3}$ in the step length function. A
realised example of the step length and turning angle relationships to revenue
can be seen at Figure \ref{S-fig:15}.

@@

We parametrised the simulation model for four populations with different
demographics; growth rates, natural mortality and recruitment functions (Table
\ref{tab:1}). Habitat preference (Figure \ref{S-fig:1}) and temperature
tolerances (Figures \ref{S-fig:3}, \ref{S-fig:4}) were unique to each
population resulting in differently weekly distribution patterns (Figures
\ref{S-fig:5}-\ref{S-fig:7}). In addition, each of the populations has two
defined spawning areas that result in the populations moving towards these
areas in pre-defined weeks (Figure \ref{S-fig:2}) with population-specific
movement rates (Table \ref{tab:1}). The individual habitat preferences and
thermal tolerances result in different spatial habitat use for each population
(Figure \ref{S-fig:9}) and consequently different seasonal exploitation
patterns (Fishing mortality in Figure \ref{S-fig:10}).

@@

With increasing probability throughout the simulation, fishing locations were
chosen based on experience of profitable catches built up in the same month
from previous years and from the previous trip. 'Profitable' in this context
was defined as the locations where the top 70 \% of expected profit would be
found given previous trips revenue and cost of movement to the new fishing
location.  This probability was based on a logistic sigmoid function with a
lower asymptote of 0 and upper asymptote of 0.95, and a growth rate that 
ensures the upper asymptote (where decisions are mainly based on past
knowledge) is reached approximately halfway through the simulation.  \\

@@

The following steps are undertaken to determine closures:
\todo{\added[id=JJ]{Procedure unclear. Refer to symbols in methods section or
		switch order starting with description of data type
		etc..}\added[id=PD]{Yes, will redo}}
\begin{enumerate}
	\item Extract data source
	\item Aggregate according to desired spatial and temporal resolution
	\item Interpolate across entire area at desired resolution using simple
		kriging using the \emph{interp} function from the R package
		akima \citep{Akima2006}.
	\item Close area covering top 5 \% of catch rates 
\end{enumerate}
In total 28 closure scenarios were run that represent combinations of:

@@

Vessels from the same fleet (and therefore targeting preference) exploit
similar but slightly different fishing grounds depending on their own personal
experience during the explore phase of the fishery (\added[id=PD]{Figure}
\ref{S-fig:14}), that is the result of the randomness in the correlated random
walk step function, with distance moved during the exploitation phase and the
direction stochastically related to the revenue experienced on the fishing
ground (\replaced[id=PD]{Figure}{while}
\ref{S-fig:15}).\todo{\added[id=JJ]{Move some of the supplementary figures to
		the manuscript}}

@@

In order to answer this question we compare different spatial and temporal
aggregations of the 'real population' distributions to:
\begin{enumerate}[label=\alph*)]
	\item \textbf{fisheries-independent data:} the inferred population from
		a fixed-site sampling survey design as commonly used for
		fisheries monitoring purposes;
	\item \textbf{fisheries-dependent data:} the inferred population from
		our fleet model that includes fishery-induced sampling
		dynamics.
\end{enumerate}

@@

The trend in fishing mortality for each species show that in most cases the
fishery closure was successful in reducing fishing mortality on the species of
interest (species 3; Figure \ref{fig:3}), though interestingly the largest
reductions in fishing mortality happened immediately after the closures,
following which the fisheries "adapted" to the closures and fishing mortality
increased again somewhat. The exception to the success was the closures
implemented based on the coarsest spatial (20 x 20) and temporal resolution
(yearly) that was ineffective with all data sources. As expected, closures
based on the "known" population distribution were most effective, with
differing degrees of success using the commercial data. Fishing mortality rates
on the other species changed in different proportions, depending on whether the
displaced fishing effort moved to areas where the populations were found in
greater or lesser density. \\

A regression tree (using the R package REEMtree \citep{Sela2012}) highlights
that the factor most contributing to differences in fishing mortality before
and after the closure was the population (72 \% showing that the closures were
effective for population 3), followed by data resolution (21 \%), data type (7
\%) with the least important factor the timescale ($<$ 1 \%). In general the
finer the spatial resolution of the data used the greater reduction in fishing
mortality for population 3 after the closures (Figure \ref{fig:4}). The notable
outliers are the commercial data at the coarsest spatial resolution (20 x 20)
at a yearly and weekly timescale, where closures were nearly as effective as
the fine-scale resolution. In this case the closures were sufficiently large to
protect a core are of the habitat for the population, but this was achieved in
a fairly crude manner by closing a large area - including area where the
species was not found (Figure \ref{S-fig:17}) that may have consequences in
terms of restricting the fishery in a much larger area than necessary. \\

@@

%% Comparison to other models
Typically, simulation models that treat fish as individuals are focussed on
exploring the inter- and intra- specific interactions among fish populations
(e.g. OSMOSE \cite{Shin2004}) in order to understand how they vary over space
and time. Our focus was on understanding the strengths and limitations of
inference from catch data obtained through commercial fishing activity with
fleets exploiting multiple fish populations and realising catch distributions
that may differ from the underlying populations. As such, we favoured a
minimum realistic model of the fish populations \citep{Plaganyi2014}, while
incorporating detailed fishing dynamics that take account of different drivers
in a mechanistic way. In this way we take account of heterogeneity in fleet
dynamics due to different preferences and drivers similarly to other approaches
\citep{Fulton2011}, but at an individual vessel rather than fleet level. We do
not explicitly define fleets as rational profit maximisers at the outset, but
consider there are several stages to development of the fishery; information
gathering through search where the resource location is not known,  followed by
individual learnt behaviour of profitable locations.  This provides a realistic
model of how fishing patterns are established and maintained to exploit an
uncertain resource through an explore-exploit strategy \citep{Mangel1983,
	Bailey2018}.

@@

In addition, we have assumed that fishing vessels are not restricted by quota
and therefore discarding of species for which vessels have no quota or that 
are unwanted is not taken into account. This is likely to be a significant
source of bias in any inference using commercial data and should also be
explored. \\
